Hydras \& Co.

### 2023-07-03 doc/chapter-fol.tex
improve chapter-fol.tex

Although all the following constructions come directly from~\cite{Goedel}, we introduced minor changes (mainly syntax) to
take into account recent changes in \coq(new constructions, tactics, notations, etc.).

@@

\subsubsection{Example: $L$, a toy language}
In order to show a few simple examples of statements and proofs, we define a small language with very few symbols:
two constant symbols: $a$ and $b$, three function symbols $f$, $g$ and $h$  (of respective arity 1, 1 and 2), three propositional symbols $A$, $B$ and $C$, a one-place predicate symbol $P$, and a binary relational symbol $R$.

@@

\begin{remark}
This representation of terms uses mutually inductive data-types instead of lists or vectors of terms. Please see also Remark~\ref{hydra:mutually-inductive-vs-lists}.
\end{remark}

\begin{remark}[Variables]
In O'Connor's formalization of first-order logic,  variables are 
just natural numbers, and the \emph{coercion} from
\texttt{nat} to \texttt{Term L} is the constructor (\texttt{@var $L$}).
Although other choices may be considered : PHOAS, de Bruijn indices, etc,  we still the data structures of~\cite{Goedel}, in order not to break long proof scripts which use this representation (please look at Section 2 of~\cite{OConnor05} for a related discussion).

@@

In~\cite{Goedel}, no \emph{constructors} of type \texttt{(Formula L)} are associated with
disjunction, conjonction, logical equivalence and existential quantifier. These constructs are formalized through \emph{definitions}\footnote{Please keep in mind that we are considering a classical logic.}.

@@

This convention allows the user to reduce to 5 (instead of 10) the number of cases in ''\texttt{match F with \dots}'' constructs. On the other hand, some computation may expand a connective like $\vee$ or
$\wedge$, or an existential quantification into a ``basic'' formula (see Sect.\vref{sect:fol-issue}).
\end{remark}

@@

For instance, the term \texttt{(var 0)}
occurs freely and also inside the scope of a quantifier in the
formula \texttt{F4} \vpageref[above]{sect:folFormExamples}.

@@

  \index{ackermann}{Projects}
  \begin{project}
 Use a HOAS representation for FOL terms and formulas (without breaking proof scripts).   
  \end{project}

@@

@@

Concerning formulas, the treatement of binding is realized 
whith the help of \texttt{List} library function \texttt{remove}.
\inputsnippets{folProp/freeVarF, FolExamples/freeVarEx}

@@

\begin{remark}
  The function \texttt{close} applies \texttt{freeVarF} and
  \texttt{List.nodup}  in order to add a sequence of universal
  quantifications \texttt{allH $i_1\, \dots\, i_k$}, in an order determined by the implementation of these functions. It may be interesting to check whether the proof of properties of \texttt{close} depend or not from this implementation.
\end{remark}
% \subsubsection{Computing a fresh variable}

@@

Let us consider for instance the formula $F = \forall\,v_1, \,\exists\,v_2,\,v_1\not=f(v_2)$ (satisfiable if we take for instance $f$ to be interpreted as the successor function).

@@

In our example, we can obtain a correct result, if
\begin{enumerate}
\item We consider a \emph{fresh} variable, \emph{i.e.}  different from $v_1$ and $v_2$, say for instance $v_3$,
\item we substitute $v_3$ to $v_2$ in $F_1$ which results in
   $F_2= \exists\,v_3,\,v_1\not=f(v_3)$
\item we substitute the term $f(v_2)$ to $v_1$ in $F_2$, which gives us 
  $\exists\,v_3,\,f(v_2)\not=f(v_3)$.
\end{enumerate}

@@

Substitution of a term to (free- occurrences of a variable 
in a formula is defined in a section of \href{../theories/html/hydras.Ackermann.folProp.html}{Ackermann.folProp}.
The definition itself takes 200 lines of \coq code, so we will only comment the \emph{structure} of this definition.

Fortunately, the reader may skip a few complex definitions 
(often because of dependent pattern matching), whose purpose
is twofold:
\begin{itemize}
\item Ensure that the substitution in a formula $F$ of a term $t$ to a variable $x$ returns a formula of the same depth as $F$,
  which allows to define a function by well-founded recursion on depth.
  \inputsnippets{folProp/substFHelp, folProp/substFDef}
\item Prove a few \emph{equations} which will be used in firther proofs on substitutions.

@@

@@

Proof trees in first-order logic are the inhabitants of the \texttt{Prf $L$} 
inductive type, defined in \href{../theories/html/hydras.Ackermann.folProof.html}{Ackermann.folProof}. Since this type has 14 constructors, we present
them one by one, after quoting its full definition.

@@

\inputsnippets{folProof/PrfDef}

@@

We want to define a data type  \texttt{Prf $l$ $F$}  of ``proof trees of the formula $F$, where $l$ is the list of assumptions used in the proof, enumerated left-to-right (\emph{i.e.} the fringe of the proof tree)''.

\inputsnippets{folProof/PrfAbout}

@@

We won't respect the order in which \texttt{Prf}'s constructor 
are enumerated in \texttt{folProof.v}. We preferred to present 
these constructors in an order inspired by a sequence
 of simple examples. On the other hand, we didn't change this order in 
\texttt{folProof.v}, in order not to break complex proofs by pattern-matching.

See \href{https://en.wikipedia.org/wiki/List_of_Hilbert_systems}.

\subsubsection{The axiom rule}

@@

\inputsnippets{folProof/AXM}

@@

\subsubsection{Modus Ponens}

@@

\inputsnippets{folProof/MP}

@@

\subsubsection{Hilbert's axioms for implication}

\inputsnippets{folProof/IMP12}

@@

\subsubsection{The rule of contraposition}

@@

\inputsnippets{folProof/CP}

@@

@@

\begin{todo}
To do propositional connectives before FO (in the doc, but not 
 in the vernacular file)
To do : EQ1 before GEN

$P,P , P\arrow P \arrow Q, \vdash Q$ ?
\end{todo}

\begin{exercise}

Prove the following lemma:
\inputsnippets{FolExamples/eqRefl}

\end{exercise}

%%%%%%%%%%%%%% New chapter 
\subsection{Contexts as sets}

\begin{exercise}

Prove the following lemma:
\inputsnippets{FolExamples/MPdiag}

\end{exercise}

\section{Derived rules and natural deduction}

 The library 
 \href{../theories/html/hydras.Ackermann.folLogic.html}{Ackermann.folLogic} contains many derived rules which allow the user to build proofs in a natural deduction style (with introduction and elimination rules).

\subsection{Example}

For instance, let us prove Peirce's rule.
 (in  ~\href{../theories/html/hydras.MoreAck.FolExamples.html}{MoreAck.FolExamples} ).

\subsubsection{Prelude}

%\inputsnippets{folPeirce/prelude}

\subsection{Proof of Peirce's law}

\inputsnippets{FolExamples/PeirceProof}

Let us try to do an implication introduction.

\inputsnippets{FolExamples/step1}

Now, we may use the law of excluded middle with the formula $P$. The only non-trivial case is about $\sim P$.

\inputsnippets{FolExamples/step2}

The rest of the proof is composed of basic proof steps, 
and bookkeeping steps (about \texttt{Ensembles.In}).

\inputsnippets{FolExamples/step3, FolExamples/step4}

\index{ackermann}{Exercises}
\begin{exercise}
Prove, using the rules described in 
 \href{../theories/html/hydras.Ackermann.folLogic.html}{Ackermann.folLogic}, the famous \emph{drinkers theorem}:

$$\exists\,x,\; (D(x)\Longrightarrow \forall\,y,\; D(y))$$
\end{exercise}

where $D$ (for ``drinks'') is some predicate symbol of arity $1$.

@@

\section{Languages for Arithmetic}

\begin{todo}
Make a chapter!
\end{todo}

Two languages built with the usual symbols of arithmetic are 
defined in ~\href{../theories/html/hydras.Ackermann.Languages.html}{Ackermann.Languages}.

\begin{itemize}
\item The first language: \texttt{LNT} (\emph{Language of Number Theory}) has just function symbols for $+$, $\times$, $0$ and successor.
\item The second language: LNN (\emph{Language of Natural Numbers})  has
the same function symbols as LNT plus one relation symbol for the strict inequality $<$ : \texttt{LT} (less than).
\end{itemize}

\subsubsection{Language of Number Theory (LNT)}

First, we declare two alphabets.

\inputsnippets{Languages/LNTDef1}

In a second time, we build \texttt{LNT} and \texttt{LNN} by filling \texttt{Language}'s \texttt{arity} field.

\inputsnippets{Languages/LNTDef2}

\begin{remark}
  We depart a little from \cite{Goedel}'s notations, where the 
function and relation symbols are called \texttt{Plus}, 
\texttt{Mult}, \texttt{LT}, etc. In our version, these type constructors are called \texttt{Plus\_}, 
\texttt{Mult\_}, \texttt{LT\_}, etc., while the names without final underscores are bound to term building functions (\emph{e.g.}
the function which takes two terms and builds the term representing their sum) (see Remark~\ref{rem:underscores}).
\end{remark}

\subsubsection{Language of Natural Numbers (LNN)}

\texttt{LNN} is an extension of \texttt{LNT}, by the addition 
of the $<$ relation symbol.

\inputsnippets{Languages/LNNDef}

\subsubsection{Examples}

Let us show a few examples (from ~\href{../theories/html/hydras.MoreAck.FolExamples.html}{MoreAck.FolExamples}).

\inputsnippets{FolExamples/arityTest}

% \subsubsection{Examples}

% With our toy language, we can build the following terms, respectively written $a$, $f(a)$ and $f(f(v_1))$ in usual mathematical notation.

% Writing these terms in bare \gallina syntax would result in quite clumsy scripts. Thus, we may use a still experimental notation scope which allows us to write/print terms and formulas in a 
% notation similar to \coq's.

% We put both the associated \coq{} terms, in bare \gallina{} syntax, and with a still experimental and provisional notation, defined in
% ~\href{../theories/html/hydras.Ackerm.FOL_notations.html}{MoreAck.FOL\_notations}).

% \inputsnippets{FolExamples/toyNotation}

% \inputsnippets{FolExamples/smallTerms}

For instance the term $v_1+0$, where $v_1$ is a variable,
is represented by the following \gallina term of type 
(\texttt{fol.Term LNT}).

\inputsnippets{FolExamples/v1Plus0}

\inputsnippets{LNN/instantiations}

\section{Notations for Formulas (experimental)}

In order to get more readable terms and formulas, we can define a few notations in ~\href{../theories/html/hydras.MoreAck.FOL_notations.html}{MoreAck.FOL\_notations} and
~\href{../theories/html/hydras.MoreAck.LNN.html}{MoreAck.LNN}.
Please note that these notation scopes are experimental: We are going to use them in examples and exercises before using them in large original proof scripts (in the \texttt{ordinals/Ackermann/} sub-directory).

We try to define notation scopes as close as possible to \coq's syntax for propositions.

Let us take for instance the following proposition (in math form):

$$\forall\, v_0,\, v_0=0\vee \exists\,v_1,\,v_1=1+v_0$$

Here is a definition, using directly the \texttt{goedel/Ackermann}'s project syntax.

\inputsnippets{LNN_Examples/uglyF0}

Note that, because of redefinitions, the disjonction \texttt{orH}
can be expanded in terms of  implication and negation (for instance when we use \texttt{Compute}).

\begin{todo}
  Present the general issue about evaluation, and our provisional solution.
\end{todo}

\inputsnippets{LNN_Examples/CNNF0}

>>>>

Although all the following constructions come directly from Russel O'Connor's work~\cite{Goedel}, we introduced minor (mainly syntactic) changes  to
take into account recent changes in \coq(new constructions, tactics, notations, etc.).

@@

\subsubsection{Example: $L$, a toy language}
In order to show a few simple examples of statements and proofs, we define a small language with very few symbols:
two constant symbols: $a$ and $b$, three function symbols $f$, $g$ and $h$  (of respective arity 1, 1 and 2), three propositional symbols $A$, $B$ and $C$, two one-place predicates symbols $P$ and $Q$, and a binary relational symbol $R$.

@@

\begin{remark}
This representation of terms uses mutually inductive data-types instead of lists or vectors of terms. Please see also Remark~\vref{hydra:mutually-inductive-vs-lists}.
\end{remark}

\begin{remark}[Variables]
In O'Connor's formalization of first-order logic,  variables are 
just natural numbers, and the conversion from
\texttt{nat} to \texttt{Term L} is the constructor (\texttt{@var $L$}).
Although other choices may be considered : PHOAS, de Bruijn indices, etc,  we still the data structures of~\cite{Goedel}, in order not to break long proof scripts which use this representation (please look at Section 2 of~\cite{OConnor05} for a related discussion).

@@

In~\cite{Goedel}, no \emph{constructors} of type \texttt{(Formula L)} are associated with
disjunction, conjonction, logical equivalence and existential quantifier. These constructs are formalized through \emph{definitions} in terms of \texttt{impH}, \texttt{notH} and \texttt{forallH}\,\footnote{Please keep in mind that we are considering a classical logic.}.

@@

This convention allows the user to reduce to 5 (instead of 10) the number of cases in ''\texttt{match F with \dots}'' terms. On the other hand, some computation may expand a connective like $\vee$ or
$\wedge$, or an existential quantification into a ``basic'' formula (see Sect.\vref{sect:fol-issue}).
\end{remark}

@@

For instance, the term \texttt{(var 0)}
occurs both freely and  inside the scope of a quantifier in the
formula \texttt{F4} \vpageref[above]{sect:folFormExamples}.

@@

  \index{ackermann}{Projects}
  \begin{project}
 Define a [P]HOAS representation for FOL terms and formulas.
Could we avoid  to  break some proof scripts?   
  \end{project}

@@

\begin{remark}
In some situations (like in the proof of \texttt{PrfEx4}~\vref{prfex4})
the user may be puzzled when a formula [s]he typed explicitely
\texttt{$\sim A$ ->$\sim B$}  will be printed 
\texttt{$A$  \textbackslash/' $\sim B$}. We will try to fix this issue.
\end{remark}

@@

Concerning formulas, the treatement of binding is realized 
whith the help of the \texttt{List} library function \texttt{remove}.
\inputsnippets{folProp/freeVarF, FolExamples/freeVarEx}

@@

\begin{remark}
  The function \texttt{close} applies \texttt{freeVarF} and
  \texttt{List.nodup}  in order to add a sequence of universal
  quantifications (\texttt{allH $i_1\, \dots\, i_k$}), in an order determined by the actual implementation of these functions. It may be interesting to check whether the proof of properties of \texttt{close} depend or not from this implementation.
\end{remark}
% \subsubsection{Computing a fresh variable}

@@

Let us consider for instance the formula $F = \forall\,v_1, \,\exists\,v_2,\,v_1\not=f(v_2)$ (satisfiable if we take for instance $f$ to be interpreted as the successor function on natural numbers).

@@

In this example, we could obtain a correct result, if
\begin{enumerate}
\item We consider a \emph{fresh} variable, \emph{i.e.}  different from $v_1$ and $v_2$, say for instance $v_3$,
\item we substitute $v_3$ to $v_2$ in $F_1$ which results in
   $F_2= \exists\,v_3,\,v_1\not=f(v_3)$
\item we substitute the term $f(v_2)$ to $v_1$ in $F_2$, which gives us 
  $\exists\,v_3,\,f(v_2)\not=f(v_3)$.
\end{enumerate}

@@

Substitution of a term to (free- occurrences of a variable 
in a formula is defined in a section of \href{../theories/html/hydras.Ackermann.folProp.html}{Ackermann.folProp}.
The definition itself takes 200 lines of \coq code, so we will only comment its structure.

Fortunately, the reader may skip a few complex definitions 
(often because of dependent pattern matching), whose purpose
is twofold:
\begin{itemize}
\item Ensure that the substitution in a formula $F$ of a term $t$ to a variable $x$ returns a formula of the same depth as $F$,
  which allows to define a function by well-founded recursion on depth.
  \inputsnippets{folProp/substFHelp, folProp/substFDef}
\item Prove a few \emph{equations} which will be used in further proofs.

@@

Let us look at a few examples. Despite the complexity of its definitoin, the function \texttt{substF} behaves well with respect with computations.

@@

Proof trees in first-order logic are the inhabitants of the \texttt{Prf $L$} 
inductive type, defined in \href{../theories/html/hydras.Ackermann.folProof.html}{Ackermann.folProof} and displayed in Figure~\vref{fig:prf-def}.

@@

\begin{figure}[p]
  \centering
\inputsnippets{folProof/PrfDef}  
  \caption{Definition of the type \texttt{Prf} of proof trees}
  \label{fig:prf-def}
\end{figure}

@@

The type  \texttt{Prf $l$ $F$}  is the type of ``proof trees of the formula $F$, where $l$ is the list of assumptions used in the proof, enumerated left-to-right (\emph{i.e.} the fringe of the proof tree)''.

In the rest of this section, we comment  every one of 
\texttt{Prf}'s 14 constructors,
 and give simple examples of their application.

@@

We won't respect the order in which \texttt{Prf}'s constructors
of type \texttt{folProof}  
are enumerated in \texttt{folProof.v} (see Figure~\ref{fig:prf-def}). 
Instead, we preferred to present 
these constructors in an order inspired by a sequence
 of simple examples. On the other hand, we didn't change this order in 
\texttt{folProof.v}, in order not to break complex proofs by pattern-matching.

For more information on Hilbert proof system, you may consult
\url{https://en.wikipedia.org/wiki/List_of_Hilbert_systems}.

\subsubsection{The axiom rule: \texttt{AXM}}

@@

@@

\subsubsection{Modus Ponens: \texttt{MP}}

@@

@@

\subsubsection{Hilbert's axioms for implication: \texttt{IMP1} and \texttt{IMP2}}

@@

\subsubsection{The rule of contraposition: \texttt{CP}}

@@

@@

\label{prfex4}

@@

\subsubsection{Rules about the universal quantifier: \texttt{FA1},
\texttt{FA2} and\texttt{FA3}}

\inputsnippets{FolExamples/PrfEx5}

\inputsnippets{FolExamples/PrfEx6}

\inputsnippets{FolExamples/PrfContrex7}

\inputsnippets{FolExamples/PrfEx8}

\subsubsection{Axioms for Equality: \texttt{EQ1} to \texttt{EQ3}}

The following proof applies \texttt{FA1} in order to build a proof
of \texttt{Prf L [] (t=t)\%fol}  for any $t$\,\footnote{Please note that the quantification on $f$ is \coq's. Thus, \texttt{eq\_ref} is kind of scheme.}

\inputsnippets{FolExamples/eqRefl}

\subsubsection{Axioms schemes for Equality: \texttt{EQ4} and \texttt{EQ5}}

The constructors \texttt{EQ4}  and \texttt{EQ5} build an infinite number of axioms, parameterized by a  relation  or function symbol.

The function \texttt{AxmEq4} generates a formula with $n$ pairs of free variables (with $n$ the arity of the considered relation symbol).

\inputsnippets{FolExamples/PrfEx910}

Please note that \texttt{EQ4} uses a \emph{sequence} of variables generated by \texttt{AxmEq4}. Any other sequence may cause 
\texttt{EQ4} to fail.

\inputsnippets{FolExamples/PrfContrex9}

The following script shows that \texttt{EQ5} is to function symbols 
what \texttt{EQ4} is to relation symbols.

\inputsnippets{FolExamples/PrfEx11}

\section{Concluding remarks}

The type \texttt{Prf} is composed of very simple rules.
Nevertheless, the examples presented in the previous section seem to show that proving even simple theorems is not trivial at all.

Indeed, in the next chapter, we will consider a proof system
\texttt{SysPrf}, based on \texttt{Prf}, the properties of which will 
allow us to prove theorems in a very simpler way.

@@
